---
title: "Skills: Correlations and model fit"
output: 
  rmdformats::readthedown:
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

A lot of quantitative analysis focuses on determining whether there is a relationship
between two variables. If two variables are related, we can use the value of
one to predict the value of the other.

For the R examples on this page, I'll start by loading the data and the packages
I need.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)

my_data <- here("datasets",
     "test.csv") |>
  read_csv()
```


A quick note about working in Excel. I'm going to put this in bold and all caps
because it's really important:

**IF YOU OPEN YOUR DATA IN A CSV FILE AND THEN ADD CALCULATIONS AND PLOTS, YOU
NEED TO SAVE YOUR WORK AS AN Excel (\*.xls or \*,xlsx`).**

# Visualizing relationships

A scatter plot is a sensible way to visualize the relationship between two 
variables. Box plots and/or violin plots offer an alternative way to visualize
the relationship between a continuous variable and a categorical variable.

## Scatter plots

Scatter plots are the most literal way to represent your data.

### R

To generate a scatter plot in R, use the `geom_point()` function. Here is how
I would show the relationship between sales price and distance to the city.

```{r}
ggplot(my_data,
       aes(x = city_dist,
           y = `sales-price`)) +
  geom_point()
```

A couple of things I notice about that plot. First of all, for both variables, 
it looks like there are a lot of lower values and fewer higher values. This 
might suggest that a log scale on both axes might make it easier to see the
relationship between these two variables.

I can transform the x or y axis by setting the `transform` parameter in the 
`scale_x_continous()` or `scale_y_continuous()` functions. While I'm at it, I 
can also name those axes to be something more descriptive.

```{r}
ggplot(my_data,
       aes(x = city_dist,
           y = `sales-price`)) +
  geom_point() +
  scale_x_continuous(name = "Distance to city (miles)",
                     transform = "log10") +
  scale_y_continuous(name = "Sales price",
                     transform = "log10") 
```

One more change I might want to make to the y-axis is to format display of the 
numbers. This is another thing I can do within the `scale_y_continuous` function.

```{r}
ggplot(my_data,
       aes(x = city_dist,
           y = `sales-price`)) +
  geom_point() +
  scale_x_continuous(name = "Distance to city (miles)",
                     transform = "log10") +
  scale_y_continuous(name = "Sales price",
                     transform = "log10",
                     breaks = c(10000,
                                100000,
                                1000000),
                     labels = c("$10,000",
                                "$100,000",
                                "$1,000,000")) 
```

Finally, this cloud of points just looks like a big black blob, with lots of 
points on top of one another. I can get a better sense of what's going on by
adjusting the point transparency. I would do this by setting the `alpha`
value within the `geom_point` function. I can also address the overplotting 
problem by making the points smaller (using the `size` parameter).

While I'm messing around the the appearance of the points, I might change the 
color as well.

```{r}
ggplot(my_data,
       aes(x = city_dist,
           y = `sales-price`)) +
  geom_point(alpha = 0.3,
             size = 0.5,
             color = "orange") +
  scale_x_continuous(name = "Distance to city (miles)",
                     transform = "log10") +
  scale_y_continuous(name = "Sales price",
                     transform = "log10",
                     breaks = c(10000,
                                100000,
                                1000000),
                     labels = c("$10,000",
                                "$100,000",
                                "$1,000,000")) 
```

### Excel

In Excel, you would create a scatter plot by selecting columns for the two variables
you want to include (hold down that ctrl or cmd key while clicking on each column
heading), and then click the picture of a little scatter plot on the Chart section
of the Insert ribbon.


```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("skills",
     "gifs",
     "scatterplot.gif") |>
  knitr::include_graphics()
```

As we say in the R example, both variables, 
have a lot of lower values and fewer higher values. This 
might suggest that a log scale on both axes might make it easier to see the
relationship between these two variables.

You can transform the axis on a scatter plot by opening the "Format axis" 
panel and checking the box for "Logarithmic scale." You might also want to 
adjust the minimum value and the value where the alternative axis crosses.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("skills",
     "gifs",
     "log-axis.gif") |>
  knitr::include_graphics()
```

You can also delete the chart title and add axis labels. To change the way
the numbers are displayed on the axis, you can just change their display in the
spreadsheet.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("skills",
     "gifs",
     "format-axis.gif") |>
  knitr::include_graphics()
```

And you can also change how the points are displayed, which could help with your
overplotting issues.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("skills",
     "gifs",
     "format-points.gif") |>
  knitr::include_graphics()
```

## Scatter plots with trend lines

If there is a weak linear relationship between your two variables, adding a 
linear trend line to your scatter plot can help emphasize that relationship.

### R

In R, you can add a linear trend line using the function `stat_smooth()`, setting
`method = "lm"`. You can also specify the color of the line and the fill color for
the confidence interval (both will be blue by default).

```{r}
ggplot(my_data,
       aes(x = city_dist,
           y = `sales-price`)) +
  geom_point(alpha = 0.3,
             size = 0.5,
             color = "orange") +
  scale_x_continuous(name = "Distance to city (miles)",
                     transform = "log10") +
  scale_y_continuous(name = "Sales price",
                     transform = "log10",
                     breaks = c(10000,
                                100000,
                                1000000),
                     labels = c("$10,000",
                                "$100,000",
                                "$1,000,000"))  +
  stat_smooth(method = "lm",
              color = "brown",
              fill = "brown")
```

### Excel

In Excel, you can you can add a linear trend line by clicking the "Add chart
element" button.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("skills",
     "gifs",
     "trendline.gif") |>
  knitr::include_graphics()
```

In that example, the trend line doesn't look like a straight line because the 
y-axis is on a log scale, and the line is referencing the underlying data rather
than the displayed data. To get to look more like a straight line, you'd actually
want to select "logarithmic" as the trend line type.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("skills",
     "gifs",
     "log-trend.gif") |>
  knitr::include_graphics()
```

Note that this trend line in the above example still doesn't look entirely 
straight. This is because setting the trend line type to "Logarithmic" will 
account for the log transformation on the x-axis, but not
on the y-axis. To account for the log-transformation on the y-axis, you'd need
to actually create a log-transformed version of your variable.

## Scatter plots categories

You can also use a scatter plot to show the relationship between a continuous
variable and a categorical variable.

### R

In R, you can use `geom_jitter` for a scatter plot where one of the values is 
categorical.

```{r}
ggplot(my_data,
       aes(x = building,
           y = `sales-price`)) +
  geom_jitter(alpha = 0.3,
             size = 0.5,
             color = "orange") +
  scale_y_continuous(name = "Sales price",
                     transform = "log10",
                     breaks = c(10000,
                                100000,
                                1000000),
                     labels = c("$10,000",
                                "$100,000",
                                "$1,000,000"))  
```


### Excel

There isn't an easy way to do this in Excel - you need to manually create a 
variable to indicate where you want the points to be positioned on the x-axis.

In this example, I'm using an if statement to set the value to 1.0 plus some 
random variation if the building category is "Single-family", to 2.5 plus some
random variation if the building category is "Duplex", and to 4.0 plus some random
variation if the building category is anything else (the remaining value is
"Three-plus").

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("skills",
     "gifs",
     "cat-vals.gif") |>
  knitr::include_graphics()
```

Then I can make a scatter plot with this new variable.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("skills",
     "gifs",
     "cat-val-plot.gif") |>
  knitr::include_graphics()
```

You'll want to turn off the x-axis in this case since the numeric values are 
meaningless and you can manually add text boxes to indicate the categories.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("skills",
     "gifs",
     "cat-val-label.gif") |>
  knitr::include_graphics()
```

## Box and whisker plots and violin plots

A couple of other plots that can summarizes differences in a continuous variable
among categories are violin plots and box-and-whisker plots (also called box
plots).

A box plot shows a rectangle representing the interquartile range for each 
category, with a bar within it indicating the median value, whiskers showing
the range of values within two interquartile ranges of the midpoint of the 
interquartile range, and points representing observations beyond that range
(outliers).

A violin plot is a nice, somewhat more intuitively interpretable alternative to
a box plot, although there isn't an easy way to produce one in Excel.

### R

In R, you can make a box plot useing the function `geom_boxplot()`.

```{r}
ggplot(my_data,
       aes(x = building,
           y = `sales-price`)) +
  geom_boxplot() +
  scale_y_continuous(name = "Sales price",
                     transform = "log10",
                     breaks = c(10000,
                                100000,
                                1000000),
                     labels = c("$10,000",
                                "$100,000",
                                "$1,000,000"))  
```

It could be helpful to overlay the box plot on a scatter plot. This functions 
like a trend line in the sense that you can see the actual data along with a
shape that emphasizes relationships that are not obvious from the scatterplot 
alone.

```{r}
ggplot(my_data,
       aes(x = building,
           y = `sales-price`)) +
  geom_jitter(alpha = 0.3,
             size = 0.5,
             color = "orange") +
  geom_boxplot(fill = NA) +
  scale_y_continuous(name = "Sales price",
                     transform = "log10",
                     breaks = c(10000,
                                100000,
                                1000000),
                     labels = c("$10,000",
                                "$100,000",
                                "$1,000,000"))  
```

A violin plot might be more intuitive than a box plot, and you can create one
using geom_violin().As a default, a violin plot is just a density curve, where 
width of the shape is proportional to the number of observations with that 
value.

```{r}
ggplot(my_data,
       aes(x = building,
           y = `sales-price`)) +
  geom_jitter(alpha = 0.3,
             size = 0.5,
             color = "orange") +
  geom_violin(fill = NA) +
  scale_y_continuous(name = "Sales price",
                     transform = "log10",
                     breaks = c(10000,
                                100000,
                                1000000),
                     labels = c("$10,000",
                                "$100,000",
                                "$1,000,000"))  
```

You can also specify that you want to add lines representing the values for 
specified quantiles. Here is how you would add lines for the 25th, 50th, and
75th percentile values.

```{r}
ggplot(my_data,
       aes(x = building,
           y = `sales-price`)) +
  geom_jitter(alpha = 0.3,
             size = 0.5,
             color = "orange") +
  geom_violin(fill = NA,
              draw_quantiles = c(0.25, 0.5, 0.75)) +
  scale_y_continuous(name = "Sales price",
                     transform = "log10",
                     breaks = c(10000,
                                100000,
                                1000000),
                     labels = c("$10,000",
                                "$100,000",
                                "$1,000,000"))  
```

### Excel

There isn't an easy way to create a violin plot in Excel, but you can create a
box-and whisker plot.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("skills",
     "gifs",
     "excel-boxplot.gif") |>
  knitr::include_graphics()
```

There is not an easy way to log-transform the numeric axis.
You would need to create a log-transformed version of the variable.



# Estimating regression models

A linear regression model will estimate the equation of a line that predicts
the value of your outcome based on the value of a predictor.

## R

In R, you estimate a linear regression model using the function `lm()`.
`summary()` will display the results of the model. 

You can choose to include only one predictor in your model. Here is how you would
estimate a model that predicts the value of sales price based on the distance to
the city.

```{r}
lm(`sales-price` ~ city_dist,
   data = my_data) |>
  summary()
```

The "Multiple R-squared" value displayed in the second-to-last line is 
the square of the correlation between the two variables.  

You can also choose to include multiple predictors in your model, including 
categorical predictors.

```{r}
lm(`sales-price` ~ city_dist + sch_quality + rooms + building,
   data = my_data) |>
  summary()
```

## Excel

In Excel, you can use the `=LINEST()` function to estimate a regression model.

You should start by setting up the table your regression results will go into.
The table should have three rows:

* Coefficient
* Standard error
* R-squared

And it should have one more column than the number of predictors you are using.
The names of the columns should be the names of your predictors, in the reverse
of the order they appear in your spreadsheet (because why not, Excel? Why not?),
and the last column should be called something like "Constant" or "Intercept".

So, something like this:

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("skills",
     "gifs",
     "regress-setup.png") |>
  knitr::include_graphics()
```

Now, select the block of cells in that table, and with all of them selected,
type the `=LINEST() function. The first argument should be your set of outcome,
the second argument should be the set of predictors, and the next two arguments
should be "TRUE" (these indicate that you want to return the standard errors and 
the R-squared value).

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("skills",
     "gifs",
     "excel-regression.gif") |>
  knitr::include_graphics()
```

The Excel regression method can only handle numeric predictors. To include a 
categorical predictor, you need to replace the categorical variable with a set 
of numeric columns of zeros and ones for each category, **except the reference
category**.
